%------------------------------------
% Dario Taraborelli
% Typesetting your academic CV in LaTeX
%
% URL: http://nitens.org/taraborelli/cvtex
% DISCLAIMER: This template is provided for free and without any guarantee 
% that it will correctly compile on your system if you have a non-standard  
% configuration.
% Some rights reserved: http://creativecommons.org/licenses/by-sa/3.0/
%------------------------------------


%!TEX TS-program = xelatex
%!TEX encoding = UTF-8 Unicode

%\documentclass[11pt, a4paper]{article}
\documentclass[11pt, letterpaper]{article}
%\documentclass[10pt, letterpaper]{article}
\usepackage{fontspec} 

% DOCUMENT LAYOUT
\usepackage{geometry} 
%\geometry{a4paper, textwidth=5.5in, textheight=8.5in, marginparsep=7pt, marginparwidth=.6in}
\geometry{letterpaper, textwidth=5.5in, textheight=9.0in, marginparsep=7pt, marginparwidth=.6in}
\setlength\parindent{0in}

% FONTS
\usepackage{xunicode}
\usepackage{xltxtra}
\defaultfontfeatures{Mapping=tex-text} % converts LaTeX specials (``quotes'' --- dashes etc.) to unicode
\setromanfont [Ligatures={Common},Numbers={OldStyle}]{Adobe Caslon Pro}
\setmonofont[Scale=0.8]{Monaco} 
\setsansfont[Scale=0.9]{Optima Regular} 
% ---- CUSTOM AMPERSAND
\newcommand{\amper}{{\fontspec[Scale=.95]{Adobe Caslon Pro}\selectfont\itshape\&}}
% ---- MARGIN YEARS
\usepackage{marginnote}
\newcommand{\years}[1]{\marginnote{\scriptsize #1}}
\renewcommand*{\raggedleftmarginnote}{}
\setlength{\marginparsep}{7pt}
\reversemarginpar

% \usepackage[misc]{ifsym} %adds email symbol ('\Letter')
\usepackage{marvosym}  %adds cell phone symbol ('\Mobilefone')
\usepackage{changepage} %allows left/right indentation of body text

% HEADINGS
\usepackage{sectsty} 
\usepackage[normalem]{ulem} 
\sectionfont{\rmfamily\mdseries\Large} 
\subsectionfont{\rmfamily\mdseries\scshape\normalsize} 
\subsubsectionfont{\rmfamily\bfseries\upshape\normalsize} 

% PDF SETUP
% ---- FILL IN HERE THE DOC TITLE AND AUTHOR
\usepackage[xetex, bookmarks, colorlinks, breaklinks, pdftitle={john_ortiz_cv},pdfauthor={John P. Ortiz}]{hyperref}  
%\hypersetup{linkcolor=blue,citecolor=blue,filecolor=black,urlcolor=blue} 
%\hypersetup{linkcolor=black,citecolor=blue,filecolor=black,urlcolor=black} 
\hypersetup{linkcolor=black,citecolor=blue,filecolor=black,urlcolor=blue} 

% define a double line separator
\def\doubleline{

	\vspace{-1.4em}
	\hspace{\fill}\linethickness{0.7pt}\line(1,0){5.5in}\hspace{\fill}
	
	\vspace{-1.0em}
	\hspace{\fill}\linethickness{0.7pt}\line(1,0){5.5in}\hspace{\fill}
	
}

% CENSORING PRIVATE INFO
% use \censor{} to black out pieces of text
% use \StopCensoring to turn off all instances of censoring
\usepackage{censor}

% MULTIPLE COLUMNS
\usepackage{multicol}
%-----------------------------------------------------------------------------------

% DOCUMENT
\begin{document}
{\LARGE John P. Ortiz} %\\[1cm]
%{\Huge John P. Ortiz} %\\[1cm]
%\vfill

%% centered info with double line underneath
%\begin{center}
%	{\LARGE John P. Ortiz} \\[0.5cm]
%	799 6th St. \#14, Los Alamos, NM 87544 •\ (541) 207-5846 •\ john.p.ortiz.14@gmail.com
%\end{center}
%\doubleline

% \StopCensoring     %<------------------ UNCOMMENT TO TURN OFF CENSORING
\begin{center}
	\begin{tabular}{l l}
		Ph.D. Candidate   &
		\hspace{.5in}\href{www.lanl.gov/EES-16}{www.lanl.gov/EES-16} \\ Johns
		Hopkins University   &
		\hspace{.5in}\href{https://johnportiz14.github.io}{johnportiz14.github.io}
		\\
		Whiting School of Engineering       & \hspace{.5in}\Email {     }
		% Whiting School of Engineering       & \hspace{.5in}
		% \href{mailto:john.p.ortiz.14@gmail.com}{john.p.ortiz.14@gmail.com}
		\href{mailto:jortiz22@jhu.edu}{jortiz22@jhu.edu}
		\\ 	
		Department of Environmental Health \& Engineering  &
		\hspace{.5in}\Mobilefone { } \censor{+1 (541) 207-5846}  \\
		% \hspace{.5in} \censor{+1 (541) 207-5846}  \\
%		Ph.D. Graduate Research Assistant    &
%		\hspace{.5in}\href{www.lanl.gov/EES-16}{www.lanl.gov/EES-16} \\
%		Los Alamos National Laboratory   &
%		\hspace{.5in}\href{http://johnportiz.weebly.com/}{www.johnportiz.weebly.com}
%		\\
%		Computational Earth Science Group (EES-16)       & \hspace{.5in}\Letter
%		{     }
%		\href{mailto:john.p.ortiz.14@gmail.com}{john.p.ortiz.14@gmail.com} \\ 
%		\censor{3215 N. Charles St. Apt. #107, Baltimore, MD 21218} &
%		\hspace{.5in}\Mobilefone { } \censor{+1 (541) 207-5846}  \\
		%\censor{5015 Carriage House,   Los Alamos, NM 87544} &
		%\hspace{.5in}\Mobilefone { } \censor{+1 (541) 207-5846}  \\		
	\end{tabular}
\end{center}

%%\hrule
%\section*{Current position}
%\emph{Emeritus Professor}, Institute for Advanced Study, Princeton

%-----------------------AREAS OF SPECIALIZATION-----------------------
%%\hrule
\section*{Areas of Specialization}
% 	numerical modeling, hydrogeology, gas flow in fractured media, petroleum
% 	geofluids
	% numerical modeling •\ hydrogeology •\ gas flow in fractured media •\
	numerical modeling •\ hydrogeology •\ fractures •\
	planetary science •\ petroleum geofluids
	

%--------------------------EDUCATION-------------------------
\section*{Education}
\noindent

\years{Current}\textsc{Johns Hopkins University (JHU)}
\begin{adjustwidth}{5pt}{0pt}
	\textsc{Ph.D.} Candidate in Environmental Health \& Engineering \
	%\textsc{Thesis:} The role of fault-zone architectural elements and basal
	%altered zones on downward pore pressure propagation and induced seismicity
	%in the crystalline basement
\end{adjustwidth}

\years{2017}\textsc{New Mexico Institute of Mining and Technology (NMT)}
\begin{adjustwidth}{5pt}{0pt}
	\textsc{M.Sc.} in Hydrology\\ %, \textsc{    GPA:} 3.89/4.00\\
	\textsc{Thesis:} The role of fault-zone architectural elements and basal
	altered zones on downward pore pressure propagation and induced seismicity
	in the crystalline basement
\end{adjustwidth}

\years{2014}\textsc{Dartmouth College}
\begin{adjustwidth}{5pt}{0pt}
	\textsc{B.A.} in Earth Sciences with Honors\\
	\textsc{Honors Thesis:} Quantifying regional sediment flux from
	observations of nearshore morphology in the Columbia River Littoral Cell 
\end{adjustwidth}


%\years{2017}\textsc{M.Sc.} in Hydrology, New Mexico Institute of Mining and Technology (NMT)
%\begin{adjustwidth}{5pt}{0pt}
%
%	\textsc{Thesis:} The Role of Fault-Zone Architectural Elements and Basal Altered Zones on Downward Pore Pressure Propagation and Induced Seismicity in the Crystalline Basement
%%	\begin{tabular}{@{}l l}
%%	\textsc{Advisor:} Dr. Mark Person & \hspace{1cm} \textsc{GPA:} 3.89/4.00 
%	 \textsc{GPA:} 3.89/4.00
%%\end{tabular}\\
%\end{adjustwidth}
%
%
%\years{2014}\textsc{B.A.} in Earth Sciences, Dartmouth College\
%\begin{adjustwidth}{5pt}{0pt}
%	\textsc{Honor's Thesis:} Quantifying regional sediment flux from observations of nearshore morphology in the Columbia River Littoral Cell 
%\end{adjustwidth}

%--------------------------PLANETARY MISSION EXPERIENCE--------------
\section*{Planetary Exploration Mission Experience}
\label{sec:missions}  %label for linking to sections
\noindent
\years{2023}TLS-SAM Experiment Proposal, Co-author
\begin{adjustwidth}{5pt}{0pt}
	Co-authored proposal with members of TLS-SAM (Tunable Laser Spectrometer
	within the Sample Analysis at Mars) team proposing strategic timing of
	atmospheric sample experiments for Mars Science Laboratory (MSL)
	\textit{Curiosity} rover to perform in order to constrain sub-diurnal
	methane variations at Gale crater, Mars. MSL \textit{Curiosity}
	successfully executed one of the proposed experiments 23 September, 2023,
	and a second on 10 December, 2023.
\end{adjustwidth}

%--------------------------HONORS/AWARDS------------------------------
%\hrule
%\section*{Grants, honors \amper{} awards}
\section*{Honors \amper{} Awards}
\label{sec:awards}  %label for linking to sections
\noindent
%\begin{adjustwidth}{5pt}{0pt}
%\end{adjustwidth}
\years{2020}R\&D 100 Award Winner (\emph{Amanzi-ATS}), Contributor
\begin{adjustwidth}{5pt}{0pt}
	Contributed to the Amanzi-ATS simulator project by augmenting code
	verification and benchmark tests in addition to maintaining user guide
	documentation. The code won an R\&D 100 Award in September 2020. 
	\href{https://www.rdworldonline.com/rd-100-award-winners-announced-in-mechanical-materials-category/}{rdworldonline.com}
\end{adjustwidth}

\years{2018}Top 20 Most Downloaded Recent Papers, Wiley Publishing (\emph{Geofluids})
\begin{adjustwidth}{5pt}{0pt}
	Amongst articles published between July 2016 and June 2018, the article
	``Exploring the potential linkages..." (see
	\hyperref[sec:pubs]{Publications \& Presentations}) was in the top 20 for
	number of downloads in the 12-months post online publication.
\end{adjustwidth}

\years{2018}“Spot" Performance Award (May), Los Alamos National Laboratory
\begin{adjustwidth}{5pt}{0pt}
	Going above and beyond the call of duty under tight and/or emergency
	deadlines to finish a project deliverable in the form of a software package
	that rapidly predicts fractured rock gas transport to the earth’s surface.
\end{adjustwidth}

%--------------------------RESEARCH GRANTS------------------------------
\section*{Research Funding}
\noindent
\emph{The asterisk (*) indicates a grant for which I was not an official PI due
to graduate student status, but on which I was the lead writer and researcher.}

\years{2021}\textsc{*CSES Student Fellowship, Planetary Science}
\begin{adjustwidth}{5pt}{0pt}
	In-depth 3-year investigation of subsurface methane transport in fractured rock
	environments on Mars. Multiple methane transport and release mechanisms 
	will be considered in addition to more accurate atmospheric forcing
	provided through subsurface-atmosphere coupling of \textsc{FEHM} subsurface
	simulator and \textsc{MarsWRF} general circulation model. 
	PI: Phil Stauffer (LANL). Co-Is: Roger Wiens (LANL), Harihar Rajaram (JHU),
	Kevin Lewis (JHU), Anthony Toigo (JHU-APL). Total: \$190k (\$126k). 
\end{adjustwidth}

\years{2021}\textsc{New Mexico Small Business Assistance (NMSBA) Program}
\begin{adjustwidth}{5pt}{0pt}
	Provided the client/small-business owner detailed workflow for analyzing
	natural fluctuations in water levels in wells in order to determine aquifer
	properties. The workflow is covered in a ~150-page user guide that provides
	detailed instructions for using several pre-existing, Open-source programs
	to perform the analysis. The workflow will allow the client to quickly and
	cheaply calculate aquifer properties as an alternative to performing
	expensive well pump tests.
	PI: Phil Stauffer (LANL). Total: \$17k (\$15k). 
\end{adjustwidth}

\years{2020}\textsc{New Mexico Small Business Assistance (NMSBA) Program}
\begin{adjustwidth}{5pt}{0pt}
	Seed funding to pursue preliminary work in tidal analysis for a client who
	owns a small environmental consulting firm. By analyzing water levels in
	wells in response to earth tides, aquifer properties can be calculated
	without the need to perform expensive pump tests.  Investigated tidal
	analysis background and methods during the first year of funding and
	delivered a 20-page report to the client on the feasibility of this method,
	highlighting a path toward a workflow that could eventually be used by the
	client.   
	PI: Phil Stauffer (LANL). Total: \$17k (\$15k). 
\end{adjustwidth}

\years{2020}\textsc{*CSES Rapid Response Research \& Development Grant, Planetary
Science}
\begin{adjustwidth}{5pt}{0pt}
	Seed funding for preliminary work and proof-of-concept results of
	subsurface methane transport in fractured rock environments on Mars
	influenced by periodic barometric pressure variations.
	PI: Dylan Harp (LANL). Co-Is: Roger Wiens (LANL), Harihar Rajaram (JHU),
	Kevin Lewis (JHU). Total: \$30k (\$24.7k). %\$30k to PI Harp.
\end{adjustwidth}


%-----------------------RESEARCH EXPERIENCE---------------------------
\section*{Research Experience}
\noindent
\years{2020 - pres.}\textsc{Doctoral Student, GRA (Graduate Research Assistant)}\\
\textit{Energy and Natural Resources Security Group (EES-16), Los Alamos National Laboratory}
\begin{adjustwidth}{5pt}{0pt}
	Carrying out programmatic work for government sponsors related to detecting
	and verifying underground nuclear explosion (UNE) tests in addition to
	performing research tasks for Ph.D. Some  continuation of research and
	responsibilities from Post-Master's Student appointment.
	
	Lead a NMSBA (New Mexico Small Business Assistance;
	\href{https://www.nmsbaprogram.org/}{https://www.nmsbaprogram.org/})
	project (2021-24) to help a small environmental consulting firm reduce
	costs associated with interpreting aquifer hydrogeologic properties from
	well data.  Demonstrated the use of frequency domain analysis and
	solid-Earth tidal loading to estimate hydrogeologic properties without the
	need for expensive pump tests. 
	% Previously Computational Earth Science.
\end{adjustwidth}

\years{2017 - 2019}\textsc{Post-Master's Student, GRA}\\
\textit{Computational Earth Science Group (EES-16), Los Alamos National Laboratory}
\begin{adjustwidth}{5pt}{0pt}
	Lead developer of the Numerical Reduced Order Multiphase Model (NROMM)
	software package for rapid prediction of gas seepage times. Software
	application is developed for use on multiple platforms
	(Windows/Linux/MacOS).\\
	Developed numerical approaches for detecting and verifying underground
	nuclear explosion (UNE) tests. Projects includes simulating
	radionuclide gas transport in fractured geologic media using finite-element
	method (FEM) and control volume finite-element (CVFEM) numerical models,
	simulating high-pressure methane injection into shale samples to inform
	laboratory investigations, and determining laboratory- and field-scale
	transport properties of rocks using models and tracer experiments.\\
	Developed the Amanzi-ATS high performance computing (HPC) flow \& transport
	simulator to meet the Nuclear Quality Assurance-1 (NQA-1) regulatory
	standard by improving code verification and benchmark tests in addition to
	maintaining software user guide documentation. Collaborated
	on a multi-lab program supported by the DOE Office of Environmental
	Management (DOE EM) to provide scientifically defensible and standardized
	assessments of the uncertainties and risks associated with the
	environmental cleanup and closure of waste sites. This software won
	an R\&D 100 Award in September 2020 (see \hyperref[sec:awards]{Honors \&
	Awards}).
\end{adjustwidth}

\years{2016 - 2017}\textsc{Graduate Research Assistant}\\
\textit{Earth and Environmental Sciences Department, NMT}
\begin{adjustwidth}{5pt}{0pt}
	Created transient 3D finite-difference (FDM) models in MODFLOW to analyze
	fluid-fault interactions as pertaining to a suite of basal reservoir
	injection scenarios. I also developed transient 2D cross-sectional FDM
	models in MATLAB to test fluid-fault interactions for crystalline basement
	fault zones exhibiting local, dynamically enhanced permeability caused by
	excess fluid pressures. In my work, I developed new approaches for
	representing fault zones with multiple architectural components and
	identified several key hydrogeologic parameters that control deep
	propagation of the fluid pressure envelope and thus present increased risk
	of induced seismic events.\\
	Collaborated on an NSF-funded (via NM EPSCoR) field project deploying
	subsurface field survey equipment (transverse electromagnetics [TEM],
	magnetotellurics [MT]) for interpretation of deep saline geothermal flow
	regimes in order to evaluate potential hydrothermal systems in southern New
	Mexico.
\end{adjustwidth}

\years{2013}\textsc{Research Intern}\\
\textit{College of Earth, Ocean, and Atmospheric Sciences, Oregon State University}
\begin{adjustwidth}{5pt}{0pt}
	Completed an NSF-funded Research Experience for Undergraduates (REU)
	internship on coastal processes by collecting and interpolating nearshore
	topographic and bathymetric data to extract key spatial and temporal
	metrics using MATLAB. Determined rates of longshore-uniform sandbar
	migration cycles along the Oregon and Washington coasts representing a huge
	component of seasonal coastal sediment flux.
\end{adjustwidth}

\years{2013}\textsc{Research Intern}\\
\textit{United States Army Corps of Engineers Field Research Facility, Duck NC}
\begin{adjustwidth}{5pt}{0pt}
	Studied coastal evolution and erosion by collecting and processing LiDAR
	observations during and after storm surges.\\
	Created a paleo-hurricane record of St. Croix by performing grain-size
	analysis on sediment cores combined with charcoal carbon dating.
\end{adjustwidth}

%-----------------------TEACHING EXPERIENCE---------------------------
\section*{Teaching Experience}
\noindent
\years{2019}\textsc{Grading Assistant}\\
\textit{Environmental Health \& Engineering Department, Johns Hopkins University}
\begin{adjustwidth}{5pt}{0pt}
	Graded weekly homework assignments for 20+ upper-level undergraduate
	engineering students in an Introduction to Fluid Mechanics course.
	Occasionally taught course lectures as needed.
\end{adjustwidth}
\years{2015 - 2016}\textsc{Graduate Teaching Assistant}\\
\textit{Earth and Environmental Sciences Department, NMT}
\begin{adjustwidth}{5pt}{0pt}
	Developed lesson plans and led class lectures and field trips for 20+ students in intermediate and upper level undergraduate Earth Sciences lab sections. Taught lab sections for courses in Geomorphology and Stratigraphy \& Paleontology with work load totaling 20 hrs/week.  
\end{adjustwidth}

%--------------------------SKILLS-------------------------------
\section*{Skills}
\subsection*{Programming/command languages}
\noindent
\textsc{Python} – High proficiency in designing and executing sensitivity
analyses of FEHM and PFLOTRAN numerical models and generating nodal grids for
finite-element meshes. Have also developed a standalone software application
for real-time predictions of gas breakthrough for use on Windows, Linux, and
MacOS platforms.\\[5pt]
\textsc{Matlab} – High proficiency in writing scripts for data processing and
developing a variety of iterative/numerical models to solve hydrogeologic
environmental problems.\\[5pt]
\textsc{Fortran} – Moderate proficiency in augmenting current and legacy flow
and transport codes (e.g., \textsc{FEHM}). Added reactive transport capability
allowing for competitive kinetic adsorption between multiple tracer species.
Maintainer of a Mars-specific branch of \textsc{FEHM}
(\textsc{FEHM-Mars}) that features code modifications to adapt to martian
conditions (e.g., reduced gravity) and a capability allowing users to specify
time-varying distribution coefficients for tracer adsorption problems.\\[5pt]
\textsc{R} – Moderate proficiency in building a variety of statistical models
to interpret and synthesize climatic, hydrologic, geochemical, and
environmental data. Collaborated on statewide New Mexico water budget research
project predicting reference evapotranspiration using different statistical
models on a wide range remote sensing parameters.\\[5pt]
\LaTeX – High proficiency in typesetting/document preparation of technical and
scientific documentation. \\[5pt]
\textsc{Bash/shell}
\subsection*{Software, codes}
\noindent
\textsc{Fehm} – High proficiency in creating multi-phase fluid flow simulations
with mass transport. Created test suites comparing vapor tracer transport
within a fracture with matrix diffusion to an analytical solution governed by a
modified Richard’s equation for unsaturated flow (with immobile liquid phase)
in order to test when multi-phase flow significantly affects gas tracer
transport. \href{https://github.com/lanl/FEHM}{github.com/lanl/FEHM} \\[5pt]
\textsc{Pflotran} – Intermediate proficiency in creating multi-phase fluid flow
simulations with mass transport. Created test suites in similar capacity to
work done with FEHM. Also used in conjunction with FEHM to verify the
capabilities of discrete fracture networks (DFNs) to simulate flow/transport in
fractured media under transient conditions.
\href{https://www.pflotran.org}{www.pflotran.org} \\[5pt]
\textsc{LaGriT} – Intermediate proficiency in generating numerical
finite-element meshes for a variety of geological applications, including
porous flow and transport modeling and discrete fracture networks. Proficient
at generating meshes using the 3 interfaces: command line, batch driven via
control file, and PyLaGrit Python command/batch interface. Have also worked on
capability development by using novel approaches to combine continuum 3D grids
with 2D discrete fracture network (DFN) planes into unified meshes for use in
transient flow and transport models.
\href{https://lagrit.lanl.gov/index.shtml}{lagrit.lanl.gov/index.shtml} \\[5pt]
\textsc{dfnWorks} – Moderate proficiency in generating 3D discrete fracture
networks (DFN) for simulating flow and transport, including generating
continuum representations of fractured porous media with octree refinement
using an upscaled discrete fracture matrix model (UDFM) workflow.
\href{https://dfnworks.lanl.gov}{dfnworks.lanl.gov} \\[5pt] 
\textsc{Modflow} – High proficiency in developing 3D transient and steady-state
FDM models to solve a variety of hydrogeologic problems including fluid-fault
interactions as related to induced seismicity in thesis work, and basin-scale
backwards modeling of groundwater solute transport as related to the Kirtland
Air Force Base spill in Albuquerque, NM.\\[5pt]
\textsc{SPOTL: Some Programs for Ocean-Tide Loading} – Moderate proficiency at
computing load tides and strains produced on the solid Earth by both ocean
tides and solid-Earth body tides.
% \url{https://igppweb.ucsd.edu/~agnew/Spotl/spotlmain.html}
\href{https://igppweb.ucsd.edu/~agnew/Spotl/spotlmain.html}{https://igppweb.ucsd.edu/~agnew/Spotl/spotlmain.html}\\[5pt]
\textsc{Sphinx} – High proficiency in generating and maintaining professional
documentation of software projects as HTML and \LaTeX \ output from
reStructuredText sources.\\[5pt]
\textsc{Git} – Moderate proficiency in source code management/version control
in application development and collaborating with colleagues on coding/software
projects. Have experience maintaining Git repositories as a lead developer as
well as in a supporting collaborator role.\\[5pt]
\textsc{ParaView} – Moderate proficiency in interactive scientific
visualization and exploration by representing data in static figures and
videos. \href{https://www.paraview.org/}{www.paraview.org} \\[5pt]
\textsc{GDB (GNU Debugger)} – Moderate proficiency with command line debugging
tool for current and legacy flow and transport codes.\\[5pt] 
\textsc{Petromod} – Moderate proficiency in creating basin evolution,
subsidence rate, and heat flow models of several American petroleum-bearing
locales.\\[5pt] 
\textsc{ArcGIS} – Moderate proficiency in using digital elevation models (DEM)
to delineate watersheds and create estimates of stream discharge in mountainous
terrain.\\[5pt] 
\textsc{Adobe Illustrator} – High proficiency in creating and editing
publication-ready figures and illustrative schematic diagrams.\\[5pt] 
\textsc{COMSOL Multiphysics} – Beginner proficiency in modeling contaminant
transport in academic coursework.\\[5pt] 
\textsc{Microsoft Excel} – Ranked \#7 in the world at Microsoft Office
Worldwide Competition 2008 in Hawaii of 56,000 initial entrants
(www.betheworldchamp.com), Ireland National Champion (of 436 entrants).
\subsection*{Other}
\noindent
Advanced proficiency in written and conversational Spanish (nine years of
academic study).


%--------------------------PUBLICATIONS--------------------------
\section*{Publications \amper{} Presentations}
\label{sec:pubs}  %label for linking to sections
% \emph{h-index:} 7 (as of August 2023)
h-index: 7 (as of August 2023)

\subsection*{Preprints}
\noindent
\years{2023}\textbf{Ortiz, J. P.}, Rajaram, H., Stauffer, P. S., 
Lewis, K. W, Wiens, R. C., Harp, D. R. 
Sub-diurnal methane variations on Mars driven by barometric pumping and
planetary boundary layer evolution. 
\emph{Authorea}. August 09, 2023. doi:10.22541/essoar.169160648.81638425/v1 \\
\emph{(Submitted to JGR: Planets)}.

%\subsection*{Journal articles \amper{} Reports}
\subsection*{Journal articles \& Reports}
\noindent
\years{2024}\textbf{Ortiz, J. P.}, Rajaram, H., Stauffer, P. S., Lewis, K. W,
Wiens, R. C., Harp, D. R.  Sub-diurnal methane variations on Mars driven by
barometric pumping and planetary boundary layer evolution.  \emph{Journal of
Geophysical Research: Planets}. doi:10.1029/2023JE008043 (in production).\\ 
\years{2022}\textbf{Ortiz, J. P.}, Rajaram, H., Stauffer, P. S., Harp, D. R.,
Wiens, R. C., Lewis, K. W.  Barometric pumping through fractured rock: a
mechanism for venting deep methane to Mars' atmosphere.  \emph{Geophysical
Research Letters}. doi:10.1029/2022GL098946. \\(\underline{Cover article}:
\href{https://agupubs.onlinelibrary.wiley.com/doi/epdf/10.1002/grl.62460}{https://agupubs.onlinelibrary.wiley.com/doi/epdf/10.1002/grl.62460}).\\ 
\years{2022}Neil, C. W., Boukhalfa, H., Xu, H., Ware, S. D., \textbf{Ortiz, J.
P.}, Avendaño, S. T., Harp, D. R., Broome, S., Hjelm, R. P., Roback, R., Brug,
W. P., Stauffer, P. H. Gas diffusion through variably-water-saturated zeolitic
tuff: implications for transport following a subsurface nuclear event.
\emph{Journal of Environmental Radioactivity}.
doi:10.1016/j.jenvrad.2022.106905.\\
\years{2021}Avendaño, S. T., Harp, D. R., Kurwadkar, S., \textbf{Ortiz, J.
P.}, Stauffer, P. H. Continental-scale geographic trends in barometric-pumping
efficiency potential: a North American case study.
\emph{Geophysical Research Letters}.  doi:10.1029/2021GL093875.\\
\years{2020}Petrie, E. S., Bradbury, K. K., Cuccio, L., Smith, K., Evans, J.
P., \textbf{Ortiz, J. P.}, Kerner, K., Person, M. A., Mozley, P. S.
Geologic characterization of nonconformities using outcrop and whole-rock core
analogues: hydrologic implications for injection-induced seismicity.
\emph{Solid Earth}, 11(5),1803-1821. doi:10.5194/se-2020-20.\\
\years{2020}Bourret, S. M., Kwicklis, E. M., Harp, D. R., \textbf{Ortiz, J.
P.}, Stauffer, P. H. Beyond Barnwell: Applying lessons learned from the
Barnwell site to other historic underground nuclear tests at Pahute Mesa to
understand radioactive gas-seepage observations. \emph{Journal of Environmental
Radioactivity}, 222. doi:10.1016/j.jenvrad.2020.106297.\\
\years{2019}Harp, D. R., \textbf{Ortiz, J. P.}, Stauffer, P. H. Identification
of dominant gas transport frequencies during barometric pumping of fractured
rock. \emph{Scientific Reports (Nature Publishing Group)}, 9(1), 9537.
doi:10.1038/s41598-019-46023-z. \\ 
\years{2019}\textbf{Ortiz, J. P.}, Person,
M. A., Mozley, P. S., Evans, J. P., Bilek, S. L. The role of fault-zone
architectural elements on pore pressure propagation and induced seismicity.
\emph{Groundwater}, 57(3): 465-478. doi:10.1111/gwat.12818\\
\years{2019}Stauffer, P. H., Rahn, T., \textbf{Ortiz, J. P.}, Salazar, L. J.,
Boukhalfa, H., Behar, H. R., Snyder, E. E. Evidence for High Rates of Gas
Transport in the Deep Subsurface. \emph{Geophysical Research Letters}.
doi:10.1029/2019GL082394.\\ 
\years{2018}Harp, D. R., \textbf{Ortiz, J. P.},
Pandey, S., Karra, S., Anderson, D., Bradley, C., Viswanathan, H., \& Stauffer,
P. H. Immobile Pore-Water Storage Enhancement and Retardation of Gas Transport
in Fractured Rock. \emph{Transport in Porous Media}, 1-26.
doi:10.1007/s11242-018-1072-8\\ 
\years{2018}Stauffer, P. H., Rahn, T. A.,
\textbf{Ortiz, J. P.}, Salazar, L. J., Boukhalfa, H., \& Snyder, E. E. Summary
of a Gas Transport Tracer Test in the Deep Cerros Del Rio Basalts, Mesita del
Buey, Los Alamos NM. United States. doi:10.2172/1417180.\\
\years{2017}\textbf{Ortiz, J. P.} The Role of Fault-Zone Architectural Elements
and Basal Altered Zones on Downward Pore Pressure Propagation and Induced
Seismicity in the Crystalline Basement. \emph{New Mexico Institute of Mining
and Technology.}\\ 
\years{2016}Zhang, Y., Edel, S. S., Pepin, J., Person, M.,
Broadhead, R., \textbf{Ortiz, J. P.}, Bilek, S. L., Mozley, P. S., \& Evans, J.
P. (2016). Exploring the potential linkages between oil-field brine
reinjection, crystalline basement permeability, and triggered seismicity for
the Dagger Draw Oil field, southeastern New Mexico, USA, using hydrologic
modeling. \emph{Geofluids}. doi:10.1111/gfl.12199\\
\years{2014}\textbf{Ortiz, J. P.} Quantifying regional sediment flux from
observations of nearshore morphology in the Columbia River Littoral Cell.
\emph{Dartmouth College Senior Honors Thesis Collection.}\\ 
\years{2014}Cohn, N., Ruggiero, P., \textbf{Ortiz, J. P.}, \& Walstra, D. J.
Investigating the role of complex sandbar morphology on nearshore
hydrodynamics. \emph{Journal of Coastal Research}, 70(sp1), 53-58.
doi:10.2112/SI65-010.1

\subsection*{Invited Talks}
\years{2023} ``The Mars Underground: Characterizing methane seepage on the Red
Planet'', LANL Center for Space and Earth Science (CSES) Symposium. 23 August,
2023.\\ 
\years{2023} ``The Mars Underground: Characterizing subsurface methane seepage
on the Red Planet'', Los Alamos Geological Society monthly meeting. 18 April,
2023.\\ 
\years{2023} ``The Mars Underground: Characterizing methane seepage on the Red
Planet'', LDRD Appraisal for FY21-23, CSES Planetary Science Student Fellow. 17 April,
2023.\\
\years{2023} ``The Mars Underground: Characterizing subsurface methane seepage
on the Red Planet'', New Mexico Bureau of Geology Seminar Series, New
Mexico Bureau of Geology and Mineralogical Resources. 7 April, 2023.\\
\years{2023} ``NROMM: Numerical Reduced-Order Multiphase Model | A tool for
making rapid predictions of UNE gas seepage'', LYNM Program Technical Meeting,
Los Alamos National Laboratory. 23 January, 2023.\\
\years{2022} ``Using legacy radionuclide data to validate barometric pumping
models'', LYNM Quad-Laboratory All-Hands Meeting, Sandia National Laboratory.
25 May, 2022.\\
\years{2022} ``PE1-A pressure and permeability predictions'', LYNM
Quad-Laboratory All-Hands Meeting, Sandia National Laboratory. 26 May, 2022.\\
\years{2022} ``Using legacy radionuclide data to validate barometric pumping
models'', LYNM Quad-Laboratory All-Hands Meeting, Sandia National Laboratory.
25 May, 2022.\\
\years{2021}
% Stauffer, P. H., Boukhalfa, H., Otto, S., \textbf{Ortiz, J. P.}, Carney, T.,
% Harp, D. R., Knight, E., Rougier, E., Rahn, T. 
``PE1-A pressure and gas arrival predictions based on in-situ permeability
measurements'', Nuclear Test Monitoring Exchange of Information by Visit and
Report (EIVR) 58. 12 October, 2021.

%\subsection*{Conference Talks \amper{} Posters}
%\subsection*{Conference Talks \& Posters}

\subsection*{Conference Talks}
\years{2023}\textbf{Ortiz, J. P.}, Rajaram, H., Stauffer, P. S., Lewis, K. W,
Wiens, R. C., Harp, D. R. Sub-diurnal methane variations on Mars driven by
barometric pumping and planetary boundary layer evolution. \textit{Session:
P44C The New Mars Underground III Oral}. AGU 2023 Fall Meeting in San
Francisco, CA.\\
\years{2018}\textbf{Ortiz, J. P.}, Harp, D. R., Stauffer, P. H. A reduced-order
model to assist real-time predictions of gas transport in unsaturated fractured
media. InterPore 10\textsuperscript{th} Annual Meeting in New Orleans, LA (May
2018).\\
\years{2018}Harp, D. R., \textbf{Ortiz, J. P.}, Stauffer, P. H., Viswanathan,
H., Anderson, D. N., Bradley, C. Analysis of enhanced gas transport in
fractured rock due to barometric pressure variations (presenting for Dylan
Harp). InterPore 10\textsuperscript{th} Annual Meeting in New Orleans, LA (May
2018).\\
\years{2017}\textbf{Ortiz, J. P.}, Ortega, A. D., Harp, D. R., Boukhalfa, H.
Improving estimates of subsurface gas transport in unsaturated fractured media
using field tracer data and numerical methods.
\textit{Session: H52A Advances in Hydrological Characterization of Flow and
Transport in Fractured Media: Numerical and Experimental Observations II}. AGU
2017 Fall Meeting in New Orleans, LA.\\
\years{2016}\textbf{Ortiz, J. P.}, Person, M. A., Mozley, P. S., Evans, J. P.
The Hydrologic Connection Between Basal Reservoir Injection, Crystalline
Basement Fault Zones, and Induced Seismicity. September 2016
GSA Annual Meeting in Denver, CO.\\

\subsection*{Conference Posters}
\years{2023}\textbf{Ortiz, J. P.}, Rajaram, H., Stauffer, P. S., Lewis, K. W,
Wiens, R. C., Harp, D. R. Sub-diurnal methane variations on Mars driven by
barometric pumping and planetary boundary layer evolution. \textit{LANL Student
Symposium}.\\
\years{2022}\textbf{Ortiz, J. P.}, Rajaram, H., Stauffer, P. H., Harp, D. R.,
Wiens, R. C., Lewis, K. W. Barometric pumping through fractured rock: A
mechanism for venting deep underground methane to Mars' atmosphere.
\textit{Session: P22F The New Mars Underground: Nexus of Decadal Planetary
Science Objectives II Poster}. AGU 2022 Fall Meeting in Chicago, IL.\\ 
\years{2018}\textbf{Ortiz, J. P.}, Harp, D. R., Stauffer, P. H., Gable, C. W.,
Makedonska, N. Coupled discrete fracture and 3D continuum domain representation
to efficiently capture gas transport from underground cavities.
\textit{Session: H51P Coupled Processes in Fractured Media Across Scales:
Experimental and Modeling Advances Posters}. AGU 2018 Fall Meeting in
Washington D.C.\\
\years{2018}Harp, D. R., \textbf{Ortiz, J. P.}, Kwicklis, E. M., Bourret, S.
M., Viswanathan, H., Stauffer, P. H. Identifying dominant barometric
frequencies driving gas transport in fractured rock 
(presenting for Dylan Harp). \textit{Session: H51P Coupled Processes in
Fractured Media Across Scales: Experimental and Modeling Advances Posters}. AGU
2018 Fall Meeting in Washington D.C.\\
\years{2013}\textbf{Ortiz, J. P.}, Ruggiero, P., Cohn, N. Interannual Sandbar
Variability within the Columbia River Littoral Cell.
\textit{Session: EP13A Coastal Geomorphology and Morphodynamics I Posters}.
AGU 2013 Fall Meeting in San Francisco, CA.

\subsection*{Departmental Seminars}
\years{2022} ``The Mars Underground: Characterizing subsurface methane seepage
on the Red Planet'', Johns Hopkins University, Baltimore, MD.
Environmental Health \& Engineering Seminar. 20 September 2022.\\
\years{2021} ``From Manhattan to Mars: Applying models of subsurface
radionuclide gas seepage from nuclear testing to understand methane release
from the martian subsurface'', Johns Hopkins University, Baltimore, MD.
Environmental Health \& Engineering Seminar. 2 November 2021.\\
\years{2021} ``Determining hydrogeologic properties using well data, barometric
pressures, and tidal analysis'', LANL, EES-16 Science Café series. 12 August 
2021.\\
\years{2021} ``From Manhattan to Mars: Generating novel insights into methane
fluctuations on the Red Planet'', Johns Hopkins University, Baltimore, MD.
Environmental Health \& Engineering Seminar. 2 March 2021.\\
\years{2019} ``Improving estimates of gas transport in fractured rock --
implications for verification of underground nuclear events'', Johns Hopkins
University, Baltimore, MD. Environmental Health \& Engineering Seminar. 29
October 2019.\\
\years{2020} ``NROMM Seepage Tool: recent capability development and analytical
verification'', LANL, EES-16 Science Café series. 23 July 2020.\\
\years{2019} ``Noble gas diffusion through variably saturated rock --
implications for verification of subsurface nuclear events'', LANL,
EES-16 Science Café series. 8 August 2019.\\
\years{2018} ``The role of fault-zone architectural elements and basement
altered zones on pore pressure propagation and induced seismicity'', LANL,
EES-16 Science Café series. 2 August 2018.\\
\years{2017} ``Improving estimates of subsurface gas transport in unsaturated
fractured media using field tracer data and numerical methods'', LANL, EES-16
Science Café series. 30 November 2017.\\



\subsection*{News/Web Articles}
% \subsection*{News/General Interest Articles}
\noindent
\years{2022}``Study illuminates Mars methane transmission from subsurface
depths that could indicate microbial source''.
\emph{STE Highlights}, August 2022. 
\href{https://www.lanl.gov/science-innovation/science-highlights/2022/2022-08.php#EarthandEnvironmentalSciences-1}
{https://www.lanl.gov/science-innovation/science-highlights/2022/2022-08.php\#EarthandEnvironmentalSciences-1}\\
\years{2022}``Study illuminates Mars methane transmission from subsurface''.
\emph{LANL News Stories}, 21 August
2022. \href{https://int.lanl.gov/news/news\_stories/2022/august/0822-mars-methane.shtml}
{https://int.lanl.gov/news/news\_stories/2022/august/0822-mars-methane.shtml}
(internal LANL webpage).\\
\years{2022}``Progress on
understanding radioactive gas migration from underground nuclear explosions''. 
\emph{STE Highlights}, February 2022.
\href{https://www.lanl.gov/science-innovation/science-highlights/2022/2022-02.php#EarthandEnvironmentalSciences-1}{https://www.lanl.gov/science-innovation/science-highlights/2022/2022-02.php\#EarthandEnvironmentalSciences-1}\\

%\years{1949}Einstein, Albert (May 1949), “Why Socialism?", \emph{Monthly Review}.
%

%--------------------------PROFESSIONAL AFFILIATIONS--------------------------
\section*{Professional Affiliations}
\years{2013 - pres.} \textsc{American Geophysical Union (AGU)}\\
\years{2015 - pres.} \textsc{American Association of Petroleum Geologists (AAPG)}\\
\years{2016 - pres.} \textsc{Geological Society of America (GSA)}\\
\years{2017 - pres.} \textsc{National Ground Water Association (NGWA)}\\
\years{2017 - pres.} \textsc{New Mexico Geological Society (NMGS)}\\
\years{2018 - pres.} \textsc{International Society for Porous Media (InterPore)}

%--------------------------SERVICE TO PROFESSION-------------------------
% or "Administrative and Collective Duties"
% - journals reviewed for, funding bodies you act as an expert for, evaluation committees on which you sit
\section*{Service to Profession}
%\subsection*{Journal Reviewer}\noindent
%\textit{Geophysical Research Letters}\\
%\textit{Hydrogeology Journal}

\subsection*{Journal Reviewer}
%\vspace{-0.9cm}

% SINGLE COLUMN------------  
	\textit{Geophysical Research Letters}\\
	\textit{Icarus}\\
	\textit{Journal of Geophysical Research: Solid Earth}\\
	\textit{Hydrogeology Journal}

% 3-COLUMN FORMAT ----------  
% \setlength\multicolsep{0pt} %reduce vertical white space at beginning of cols
% % ROW1 --------------------  
% \begin{multicols}{3}
	% % \textit{J. of Geophysical Research}\\
	% \textit{JGR: Planets}\\
	% \textit{Geophysical Research Letters}\\
	% \textit{Hydrogeology Journal}
% \end{multicols}
% % ROW2 --------------------  
% \begin{multicols}{3}
	% \textit{Icarus}\\
	% % \textit{BLANK1}\\
	% % \textit{BLANK2}
% \end{multicols}

\subsection*{Advisory Roles}\noindent

\years{2018 - 2019} \textit{LANL Student Programs Advisory Committee (SPAC)}\\
Served as a 1-year appointment as Graduate Student Representative to a committee that advises Los Alamos National Laboratory's Student Programs Office on matters related to the hiring and general well-being of student employees.



%\vspace{1cm}
\vfill{}
%\hrulefill

\begin{center}
{\scriptsize  Last updated: \today\- •\- Typeset in \href{http://nitens.org/taraborelli/cvtex}{
\fontspec{Times New Roman}\XeTeX }}
\end{center}
% ---- FILL IN THE FULL URL TO YOUR CV HERE
%\href{http://nitens.org/taraborelli/cvtex}{http://nitens.org/taraborelli/cvtex}}
%\end{center}

\end{document}
